DOpElib:\\ Differential Equations and Optimization Environment

### 2013-12-13 Examples/PDE/InstatPDE/Example1/content.tex
PDE/InstatPDE Example 1, 3, 4, 5 ueberarbeitet

\subsubsection{General problem description}
In this example we consider the nonstationary incompressible Navier-Stokes equation. As in the stationary PDE Example 3 \ref{PDE_Stat_FSI_STVK},
we use the symmetric fluid stress tensor, i.e. 
in strong formulation we deal with 
\begin{align*}
\rho \partial_t v - \rho \nabla\cdot (\nabla v + \nabla v^{T}) + \rho (v\cdot\nabla)v  +\nabla p &= f \\
\nabla\cdot v &= 0 
\end{align*}
on time interval $I=[0,T]$ and the fluid benchmark domain (Schaefer/Turek 1996) 
Here, we set $f=0$ and the flow is driven by a Dirichlet inflow condition.

@@

\begin{align*}
v^{n+1} - k\theta \left(\nabla\cdot (\nabla v^{n+1} + \nabla {v^{n+1}}^T) + \nabla p^{n+1}\right) &=
k\theta f^{n+1} + k(1-\theta)f^{n}\\
&+  v^n + k(1-\theta) \left(\nabla\cdot (\nabla v^n + \nabla (v^n)^{T}) 
+ \nabla p^n\right) \\
\nabla \cdot v^{n+1} &= 0 
\end{align*}
In the case of the BE-scheme, $\theta = 1$, and the equation is reduced to
\begin{align*}
v^{n+1} - k (\nabla\cdot (\nabla v^{n+1} + \nabla {v^{n+1}}^{T}) + \nabla p^n) &=
k\theta f^{n+1} +  v^n  \\
\nabla \cdot v^{n+1} &= 0
\end{align*}
Note, that one should prefer a complete implicit treatment of the
pressure $p$. Instead of using $\theta p^{n+1} + (1-\theta)p^n$, the pressure
appears only with $\theta p^{n+1}$.

@@

\begin{Problem}[Backward Euler (BE) timestepping problem]
Let $\theta =1$.Find $v:= v^{n+1}\in V$ and $p:= p^{n+1}\in L$:
\begin{align}
(v,\phi^v) + k\theta (\nabla v + \nabla v^{T}, \nabla\phi^v) 
- k  (p, \nabla\cdot\phi^v) &=
k\theta (f^{n+1},\phi^v) +  (v^n,\phi^v)\\
(\nabla \cdot v,\phi^p) = 0  \label{eq.divfreiheit}
\end{align}
\end{Problem}
for all suitable test functions ${\phi^v , \phi^p} \in V\times L$.

>>>>

\subsubsection{General problem description}
In this example we consider the nonstationary incompressible Navier-Stokes equation. As in the stationary PDE Example \ref{PDE_Stat_Stokes},
we use the symmetric fluid stress tensor, i.e. 
in strong formulation we deal with 
\begin{align*}
\rho \partial_t v - \rho \nabla\cdot (\nabla v + \nabla v^{T}) + \rho (v\cdot\nabla)v  +\nabla p &= f \\
\nabla\cdot v &= 0 
\end{align*}
on time interval $I=[0,T]$ (with $T=80$) and the fluid benchmark domain (Schaefer/Turek 1996). 
Here, we set $f=0$ and the flow is driven by a Dirichlet inflow condition.

@@

\begin{align*}
v^{n+1} - k\theta \left(\nabla\cdot (\nabla v^{n+1} + \nabla {v^{n+1}}^T) + \nabla p^{n+1}\right) =&
k\theta f^{n+1} + k(1-\theta)f^{n}\\
&+  v^n + k(1-\theta) \left(\nabla\cdot (\nabla v^n + \nabla (v^n)^{T}) 
+ \nabla p^n\right) \\
\nabla \cdot v^{n+1} &= 0 
\end{align*}
In the case of the BE-scheme, $\theta = 1$, and the equation is reduced to
\begin{align*}
v^{n+1} - k (\nabla\cdot (\nabla v^{n+1} + \nabla {v^{n+1}}^{T}) + \nabla p^n) &=
kf^{n+1} +  v^n  \\
\nabla \cdot v^{n+1} &= 0
\end{align*}
Note, that one should prefer a complete implicit treatment of the
pressure $p$. Instead of using $\theta p^{n+1} + (1-\theta)p^n$, the pressure
appears only with $\theta p^{n+1}$.

@@

\begin{Problem}[Backward Euler (BE) timestepping problem]
Let $\theta =1$. Find $v:= v^{n+1}\in V$ and $p:= p^{n+1}\in L$:
\begin{align}
(v,\phi^v) + k\theta (\nabla v + \nabla v^{T}, \nabla\phi^v) 
- k  (p, \nabla\cdot\phi^v) &=
k\theta (f^{n+1},\phi^v) +  (v^n,\phi^v)\\
(\nabla \cdot v,\phi^p) = 0  \label{eq.divfreiheit}
\end{align}
\end{Problem}
for all suitable test functions ${\phi^v , \phi^p} \in V\times L$.

### 2013-12-13 Examples/PDE/InstatPDE/Example3/content.tex
PDE/InstatPDE Example 1, 3, 4, 5 ueberarbeitet

%\subsubsection{Program description}

>>>>

\subsubsection{Program description}
Note that the initial conditions are only $H^1$-regular. Because of this, the Crank-Nicolson scheme, which is not strongly A-stable, is not suited to solve this problem. As we want to use a second order accurate time stepping scheme, we use the shifted Crank-Nicolson method. The rest of the program is as before.

### 2013-12-13 Examples/PDE/InstatPDE/Example4/content.tex
1) Minor changes in the formatting of the documentation. 2) Added documentation to PDE/StatPDE/Example9
Removed Typos in Documentation; OPT/StatPDE/Example1 documentation has been augmented; Introduction has been augmented
PDE/InstatPDE Example 1, 3, 4, 5 ueberarbeitet

\textbf{General problem description}

\vspace{0.2cm}

In this example we consider one of the prototypical instationary equations, the parabolic heat equation
\begin{align*}
\partial_t u(t,x) - \Delta u(t,x) &= f(t,x),\\
u(t,x)|_{\partial \Omega} &= g(t,x),\\
u(0,x) &= u_0(x)
\end{align*}
with unknown solution $u:\Omega \rightarrow \mathbb{R}$, where $\Omega \subset \mathbb{R}^d$. In our example, we consider the simplest case $d=1$, where the Laplacian $\Delta$ reduces to $\partial_x^2$. The computational domain is $\Omega \times I = [0,1] \times [0,1]$. For further simplification, we choose the righthand side as $f=0$ as well as homogeneous Dirichlet boundary conditions ($g=0$). The initial condition is given by $u_0(x) = \min(x,1-x)$.
\vspace{0.2cm}

\textbf{Program description}

\vspace{0.2cm}

There are few new things compared to the first two nonstationary examples. This is the first time we solve an equation in one spatial dimension. In most cases, the dimension dependence is covered by the \texttt{LOCALDOPEDIM} and \texttt{LOCALDEALDIM} variables (which are defined at the beginning of the \textit{main.cc} file), but there might be some places in the code (especially your own code) where a concrete dimension number is given to an object. There you have to replace it manually. Do not forget to insert the correct dimension in the \textit{Makefile}!\\
The most important feature of this example is the serial application of several time-stepping schemes. At the moment, the following schemes are available (see also example \ref{PDE_Instat_Stokes}):
\begin{enumerate}
\item
Forward Euler scheme (FE)
\item
Backward Euler scheme (BE)
\item
Crank-Nicolson scheme (CN)
\item
shifted Crank-Nicolson scheme (sCN)
\item
Fractional-Step-$\theta$ scheme (FS)
\end{enumerate}
All these time-stepping methods are applied in the current example in order to check them and to compare their characteristics. To keep the computing time acceptable, we choose a one dimensional example.\\
One more innovation is the output format. We want to represent the output at single timepoints as a function graph on the space interval $[0,1]$; this can be done using \textit{GNUPLOT}, for example, so instead of \textit{.vtk} files as in all former examples, we now write out \textit{.gpl} files.

>>>>

\subsubsection{General problem description}

In this example we consider one of the prototypical instationary equations, the parabolic heat equation, i.e. for $x \in \Omega \subset \mathbb{R}^d, t\in I =[0,T], T\in \mathbb{R}^+,$ we search for the unknown solution $u:I \times \Omega \rightarrow \mathbb{R}$
\begin{align*}
\partial_t u(t,x) - \Delta u(t,x) &= f(t,x),\\
u(t,x)|_{\partial \Omega} &= g(t,x),\\
u(0,x) &= u_0(x).
\end{align*} In our example, we consider the simplest case $d=1$, where the Laplacian $\Delta$ reduces to $\partial_x^2$. The computational domain is $ I\times \Omega = [0,1] \times [0,1]$. For further simplification, we choose the right hand side as $f=0$ as well as homogeneous Dirichlet boundary conditions ($g=0$). The initial condition is given by $u_0(x) = \min(x,1-x)$.

\subsubsection{Program description}

There are few new things compared to the other nonstationary examples. This is the first time we solve an equation in one spatial dimension. In most cases, the dimension dependence is covered by the \texttt{LOCALDOPEDIM} and \texttt{LOCALDEALDIM} variables (which are defined at the beginning of the \textit{main.cc} file), but there might be some places in the code (especially your own code) where a concrete dimension number is given to an object. There you have to replace it manually. Do not forget to insert the correct dimension in the \textit{Makefile}!\\
The most important feature of this example is the serial application of several time-stepping schemes. At the moment, the following schemes are available (see also example \ref{PDE_Instat_Stokes}):
\begin{enumerate}
\item
Forward Euler scheme (FE)
\item
Backward Euler scheme (BE)
\item
Crank-Nicolson scheme (CN)
\item
shifted Crank-Nicolson scheme (sCN)
\item
Fractional-Step-$\theta$ scheme (FS)
\end{enumerate}
All these time-stepping methods are applied in the current example in order to check them and to compare their characteristics. To keep the computing time acceptable, we choose a one dimensional example.\\
One more innovation is the output format. We want to represent the output at single timepoints as a function graph on the space interval $[0,1]$; this can be done using \textit{GNUPLOT}, for example, so instead of \textit{.vtk} files as in all former examples, we now write out \textit{.gpl} files.
